Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-03-14 ribose-paper.RMD
updated discussion, needs more work

The main findings of the present study were that resistance training with glucose for five sessions did not affect markers of ribosome biogenesis, measured as total RNA, rRNA and rDNA transcription initiation associated proteins, compared to RT with placebo. Similarly, RT with glucose did not affect markers of skeletal muscle functionality such as muscular performance and -recovery, total session volume, and training intensity, compared to RT with placebo. There are studies suggesting that high vs. low glucose treatment/glucose starvation impacts ribosome biogenesis in yeast, rodent and human non-muscle cells (KIM13;MARIAPPAN11;TANAKA15;ZHAI12), however, this effect does not seem evident when total daily glucose intake is matched.

To our knowledge, no previous study has investigated the effects of glucose compared to placebo ingestion with RT on ribosome biogenesis in human skeletal muscle cells, making our hypotheses exploratory and comparisons not straight forwards. The current model of ribosome biogenesis suggests that changes in the accumulation of total RNA and rRNA early after the onset of an RT period predict long-term RT adaptations (FIGUEIREDO19; HAMMARSTROM20; HAMMARSTROM21).Therefore, we assumed 2 weeks to be a suitable time frame for a moderate RT volume to elicit changes in ribosome biogenesis, as similar observations have been made previously (FIGUEIREDO16; HAMMARSTROM20; HAMMARSTROM21). In Hammarström et al. (2020), participants that increased ribosome biogenesis after two weeks (5 RT sessions) of RT, also responded well after the entire intervention (12 weeks, 31 sessions). Figueiredo and colleagues (2016) also observed markedly acute effects, where 47S pre-rRNA increased after a single RT session, passing baseline after 4hrs, peaking at 24hrs and remaining elevated until 48hrs post-exercise. The four mature rRNA did not measurably increase until after multiple sessions (FIGUEIREDO16). As such, increases in total RNA and rRNA observed in the present study are in line with previous observations (FIGUEIREDO16; HAMMARSTROM20; H21?), and also evident i moderately trained individuals, compared to the untrained or RT-untrained individuals of previous studies (FIGUEIREDO16;HAMMARSTROM20;H21?).

Further, the accumulation of total RNA and rRNA is also suggested to reach a plateau after approximately 5 RT-sessions, thus, in addition to the demanding RT- and dietary protocol and the exploratory nature of the hypotheses, we did not conduct a more long-term intervention. COMMENT ON PROTEIN DATA HERE?

As we did not observe any effect of ingesting glucose before and after RT compared to placebo in this design, there is little to suggest that glucose *per se* has any additional effect on ribosome biogenesis, as long as the daily energy intake is sufficient to support cellular energy levels required to drive this process. Looking at previous observations from *in vitro* studies of rodents, yeast, and non-muscle cell cultures (MARIAPPAN11; KIM13; TANAKA18; ZHAI12; HOPPE09; MURAYAMA08), one interesting notion would be how high glucose vs. low glucose or glucose starvation would affect ribosome biogenesis over time in human skeletal muscle cells. Time being of the essence in such a perspective, as previous *in vitro* studies have utilized cells with higher physiological rate than human skeletal muscle cells (KIM13;MARIAPPAN11;TANAKA15;ZHAI12).

Indeed, our findings suggest that glucose ingestion *per se* before and after RT does not induce changes in the internal physiological milieu of human skeletal muscle cells with anabolic effects. It has been theorised that glucose may convey an anabolic signal through insulin, and thus operate as a signal mediator affecting upstream pathways of ribosome biogenesis (KUSNADI15; TANAKA18). Such an anabolic signal could then possibly increase PIC assembly, and consequently Pol I recruitment, initiating rDNA transcription and subsequent rRNA maturation, leading to increased ribosome content (FIGUEIREDO19; TANAKA18). As significantly higher blood levels of c-peptide and glucose were observed in RT with glucose compared to placebo in the present study, without any difference in ribosome biogenesis, there is little to suggest that either insulin or glucose has any direct acute effects.

Another theory is that glucose ingestion leads to higher cellular energy levels compared to placebo, stimulating energy-sensitive pathways such as PIH1, mTORC1, ERK1/2, AMPK and SIRT1 in a dose-response relationship (KIM13;MARIAPPAN11;TANAKA15;ZHAI12). The mTORC1, along with ERK1/2 and PIH1 up-regulates PIC assembly during high-energy conditions, while AMPK and possibly SIRT1 represses PIC assembly during low-energy conditions, as observed in rodent (KILDE), yeast (KILDE), and non-muscle cells (KILDE) *in vitro*. Specifically, high glucose treatment augmented rDNA transcription by increased UBF and SL-1 activity, while low glucose treatment, glucose starvation, or nutrient depletion stimulated AMPK-dependent TIF-IA and chromatin regulation suppression (KOS-BRAUN17; MARIAPPAN11; ZHAI12). In addition to UBF and SL-1, assembly of the PIC requires TIF-IA and Pol I recruitment, induced by amino acid intake in an mTOR-S6K-dependent manner and possibly by direct c-Myc activation (KUSNADI15). As such, increasing energy supply and nutrient availability should in line with these theories potently stimulate PIC assembly and rDNA transcription. Importantly, these pathways were not directly investigated in the present study. However, downstream targets (UBF, c-Myc, rpS6?) remained unaffected by glucose compared to placebo ingestion, suggesting there is not a dose-response relationship between muscle cellular energy levels and ribosome biogenesis during an RT-intervention (Not sure if I can say this).

A weakness in the present study, is the lack of data on intracellular glucose levels in the muscle cells. Thus, it cannot be said for sure whether there were differences between glucose and placebo or not.

A possible explanation for the lack of effect of glucose compared to placebo, is that both treatments may have exercised with sufficient levels of nutrients and energy, meaning that mTORC1 had sufficient nutrient availability and, thus was not inhibited by AMPK due to energy deficiency (HOPPE09; KUSNADI15). Indeed, mTORC1 is characterised as energy-sensitive (DENNIS01), and therefore it might have been directly affected by high vs. low-glucose conditions (MARIAPPAN11). However, previous studies suggest that mTORC1 is not sensitive enough to register levels of ATP, rather depending on signals from AMPK to assess ATP levels (KIM10). This would arguably make AMPK the primary regulator of the potential impact of glucose on ribosome biogenesis, and thus, glucose has no direct effect on mTORC1 signalling (REFERENCES?). Importantly, intramuscular glycogen content was not measured in the present study, therefore we cannot determine whether cellular energy levels in fact were equal or not, or if there is reason to assume that AMPK-dependent mTORC1 signalling was different in the two treatments. Potentially, such data could give higher resolution to our data. Along with the robust increases in proxy markers of ribosome biogenesis, the equal macronutrient intake between treatments on pairwise consecutive days and plasma glucose levels not indicating energy deficiency during RT, potent mTORC1 stimulation without AMPK-dependent inhibition may be a plausible explanation. Together with previous studies on this topic (MARIAPPAN11; ZHAI12), there are currently scarce observations to suggest glucose carries a direct anabolic signal. As such, physiological levels of plasma glucose during RT, combined with a sufficient daily macronutrient intake, seem potent enough to supply the energy needed for ribosome biogenesis. The lacking differences between treatments in the biological data are further supported by our observations on functional data.

Overall, glucose did not affect the isometric or isokinetic peak torque at 240 d/s, or muscular recovery following the sixth and last RT session, measured as temporal changes in isometric and isokinetic peak torque. Generally, the mean isometric and isokinetic peak torque slightly decreased during the intervention, especially between session 4 and 6, and regained after the intervention. Further, there were no differences between glucose and placebo in total session volume or training intensity. Both exercise with glucose and placebo yielded higher total session volumes at higher percentage of 1RM at the last session compared to baseline, as expected from earlier reports on the moderate-volume RT protocol (KRIEGER09; RHEA03; SCHOENFELD17), in addition to a more recent study implementing the unilateral protocol (HAMMARSTROM20+21/22). Interestingly, we did observe a significant difference between treatments in mean change of isokinetic peak torque at 60 d/s before the sixth RT session, where exercise with glucose decreased peak torque 9% less compared to placebo. However, as none of the other measurements indicated any difference between treatments, a true effect of glucose seems unlikely. Taken together, these results may add to previous observations with no differences between co-ingestion of carbohydrate and protein compared to protein alone, in regard to skeletal muscular recovery following RT (KOOPMAN07; STAPLES11). Collectively, the functional data point towards an exhaustive, yet growth-inducing RT protocol, especially considering the biological data. As a side note, an interesting observation was made by another master’s thesis on this project, finding significantly lower MuRF protein levels at post with glucose treatment compared to placebo. This may suggest an effect of glucose on protein degradation (WILKES09), however, whether this contributes to growth or simply protein turnover has been a topic of debate (FIGUEIREDO19).

Despite being a thoroughly standardised design, a limitation may be posed through lack of participants. We initially reach our minimum limit for statistical power, i.e. 16 participants, however, three dropped out during the intervention, possibly impacting our power. Coupled with the fact that no previous study could guide our power calculations, this, of course, introduces some degree of uncertainty regarding our results. However, we’ve reported little to suggest any true acute effect of RT with glucose ingestion compared to placebo. Moreover, the study was specifically designed to investigate differences in pre-post changes within participants, and thus also quite limited to this perspective. Hence, the effects of time observed exclusively serve as supplementary to the efficiency of the design, without a negative control group to either confirm or refute them. If further investigations are to be done on this topic, these analyses should include more participants and optimally a measure of intramuscular glycogen levels.

In conclusion, resistance exercise training combined with glucose ingestion does not acutely augment ribosome biogenesis following 2 weeks of moderate-volume resistance training in moderately trained young male and female adults. Neither does it acutely affect muscular performance in isometric and isokinetic knee extension peak torque, nor enhance post resistance training recovery within 24hrs.

>>>>

The main findings of the present study were that heavy resistance training with glucose did not affect markers of ribosome biogenesis, measured as total RNA, rRNA and rDNA transcription initiation associated protein, compared to RT with placebo. Similarly, RT with glucose in general did not affect markers of skeletal muscle functionality such as strength and -recovery, or total training session volume. At one time point, RT with glucose led to less reduction in strength compared to placebo, and levels of plasma glucose and c-peptide were significantly higher before, during and after the RT sessions with glucose. Further, there were no differences in daily macronutrient intake on pairwise consecutive days. This would suggest the effects of high vs. low glucose/glucose starvation treatment on ribosome biogenesis or rDNA transcription initiation observed in previous in vitro studies [@mariappan11; @tanaka15] are not translatable to acute effects in human skeletal muscle in vivo.

To our knowledge, no previous study has investigated the effects of glucose compared to placebo ingestion with RT on ribosome biogenesis in human skeletal muscle, making our hypotheses exploratory and comparisons not straight forwards. Mariappan and colleagues (2011) investigated the effect of high glucose treatment on ribosome biogenesis in glomerular epithelial cells (GECs) from diabetic (type 1 and 2) mice, observing a UBF-dependent augmentation of ribosome biogenesis. In a study from 2015, Tanaka et al. (2015) observed that mild glucose starvation reduced rRNA transcription in human breast cancer cells. These findings do imply an important role for glucose in the initiation of rDNA transcription, the rate-limiting step of ribosome biogenesis [@mariappan13; @tanaka15]. In the present study, we did not find such indications of glucose having an effect as these previous studies. Notably, kidney cells from mice and human breast cancer cells are likely to have a higher physiological rate compared to human skeletal muscle cells, thus, our two-week intervention may not be directly comparable to these findings, as was expected. Nevertheless, we did observe robust increases in both total RNA and 47S pre-rRNA and all four mature rRNAs (18S, 28S, 5.8S, 5S) after five RT sessions with both treatments. This means that the time frame and training volume chosen for this design was adequate to elicit changes in rDNA transcription and subsequently ribosome biogenesis by proxy. Further, the training, testing and biological sampling were conducted under standardized conditions, and there were no differences between pairwise consecutive days in macronutrient ingestion and training volume. As such, the day-to-day and test-to-test variation should have been minimal, especially when eliminating between-subjects biological variation with a within-subjects design as well (SOURCE). Importantly, both Mariappan (11) and Tanaka (15) compared high glucose treatment to either low glucose treatment or glucose starvation. In the present study however, we compared RT with either protein and glucose or protein and placebo. The participants fasted from 2200hrs the preceding day until the morning of each training day, receiving only the prescribed treatment until 2.5 hrs after RT. Thus, except from approximately 0700hrs-1230hrs, the participants ingested an equal amount of carbohydrates on pairwise days. As such, there were no differences in baseline levels of plasma glucose prior to the daily intervention, as by design, since we were interested in investigating if there were direct effects of glucose as a signalling molecule, based on the observations of Tanaka et al. [@tanaka2015] and Mariappan et al. [@mariappan2011]. Further, the significant difference in levels of plasma glucose and c-peptide measured before, during and after session six, confirms that the treatment during of RT with glucose indeed was sufficient. Despite of this, we found no difference in markers of ribosome biogenesis between RT with glucose and RT with placebo, this does suggest that there are no additional effects of exercising with glucose, when total macronutrient intake is standardized.

Prior to the present study, it was theorized that glucose may convey an anabolic signal, either directly or through insulin, and thus operate as a signal mediator affecting upstream pathways of ribosome biogenesis [@kusnadi15; @tanaka18]. This theoretical anabolic signal could possibly stimulate PIC assembly and Pol I recruitment to the rDNA promoter, initiating rDNA transcription and subsequent rRNA maturation, and lastly increased ribosome content of the skeletal muscle cell [@figueiredo19; @tanaka18]. Thus, we hypothesized that if glucose directly or indirectly via insulin can stimulate initiation of rDNA transcription, five RT sessions with glucose will elicit this effect compared to five RT sessions with placebo.  This time frame has previously shown to be sufficient to measure changes in markers of ribosome biogenesis [@hammarstrom20; @hammarstrom21; @figueiredo16]. <!--double-check these references--> As no such effect was observable in the present study, this hypothesis was rejected. Further, glucose ingestion affects cellular energy levels; thus another theory is that high glucose treatment would stimulate energy-sensitive pathways such as PIH1, mTORC1, ERK1/2, AMPK and SIRT1 in a dose-response relationship [@kim13; @mariappan11; @tanaka15; @zhai12]. <!-- check if this is right --> Mammalian target of rapamycin complex 1, along with ERK1/2 and PIH1 upregulates PIC assembly during high-energy conditions, while AMPK and possibly SIRT1 represses assembly of the PIC during low-energy conditions, as observed in rodents [SOURCE], yeast [SOURCE], and human non-muscle cells [SOURCE] *in vitro*. Specifically, high glucose treatment augmented rDNA transcription by increased UBF and SL-1 activity, while low glucose treatment, glucose starvation, or nutrient depletion stimulated AMPK-dependent TIF-1A and chromatin regulation suppression [@kos17: @mariappan11; @zhai12]. <!--check these references--> In addition to UBF and SL-1, assembly of the PIC requires TIF-1A and Pol I recruitment, induced by amino acid intake in an mTOR-S6K-dependent manner and possibly by direct c-Myc activation [@kusnadi15]. As such, increasing energy supply by RT with glucose vs. RT with placebo and nutrient availability should in line with these observations potently stimulate PIC assembly and rDNA transcription. Though we did not investigate these pathways directly, the downstream targets UBF and rpS6 did not differ in mean change between treatments. As the participants in the present study fasted until the daily training was completed, and matched their daily energy intake on pairwise days, this suggests that as long as the daily energy requirements are met, ingesting glucose with RT does not stimulate further gains in anabolic responses. Notably, the daily RT protocol was quite short (~30min), which may have reduced the effectiveness of the energy addition through glucose. In this regard, it would have been interesting to measure the intramuscular glycogen storages, to see how the RT sessions impacted intracellular energy levels. Unfortunately, such analyses were not conducted in the present study. Regrettably, this is a weakness of the design, as we have not measured directly what happened to the excess plasma glucose. Nevertheless, the absent effect of increasing energy availability during RT suggest that, there is no dose-response relationship between acutely increasing myocellular energy levels before and after RT and ribosome biogenesis.

Interestingly, we did observe a linear relationship between total RNA and UBF, suggesting ribosome biogenesis to be dependent on UBF activation, in line with observations in Mariappan et al. (2011). This was not evident with total RNA and c-Myc or rpS6, which we observed no linear relationship between. The upstream binding factor (UBF) has indeed previously been described as a master regulator of rDNA transcription, based on findings from cell cultures [@figueiredo19; @kusnadi15; russel&zomerdijk2005]. However, it has only recently been found in a linear relationship with total RNA content in human skeletal muscle cells after a short period of RT [@hammarstrom21], which the observation of the present study supports. As such, UBF also seem to respond to mechanical loading in human muscle cells in line with responses seen in cell cultures and synergist ablation models [@mariappan11; @vonwalden18], in addition to the previously reported response to hormonal, nutritional, and cellular energy signals [@figueiredo19; @kusnadi15; russel&zomerdijk2005]. While c-Myc and rpS6 did not increase linearly with accumulation of total RNA, the cellular content of c-Myc and rpS6 increased from baseline to post with both treatments. The transcription initiation factor c-Myc has been implicated as important to ribosome biogenesis [@vonwalden19; @figueiredo19], though its role is likely as a general transcription factor rather than specific, compared to UBF [SOURCES?]. Further, the ribosomal protein S6 (rpS6) has previously been shown to correlate strongly with rRNA content after synergist ablation [@nakada16; @chaillou], <!--check Nakada 16 for these refs--> and is suggested as a reliable measurement of ribosome biogenesis [@nakada16]. Together, this may explain why the between treatment difference in c-Myc content did not induce differences in total RNA or rRNA content, as the specific transcription factor UBF and the rpS6 increased equally with both treatments. Notably, the difference between c-Myc in glucose and placebo may have caused responses in pathways not measured in the present study.

The results from the strength tests do indicate that the participants might have been fatigued during the intervention, as they exercised every day for 12 days straight, without breakfast other than the treatments supplied. In this perspective, we expected glucose to increase energy availability and thus, muscle performance during strength testing.  However, this was not evident, as the mean response in strength as an index of peak torque at all speeds decreased with both treatments from baseline to post five and six sessions. Notably, RT with glucose reduced strength significantly less than RT with placebo post five sessions, which may indeed have been an effect of increased energy availability, seeing as glucose is the preferred energy source during strenuous activity [@mul15]. <!--check this ref--> Arguably, this might imply that RT with placebo experienced a higher stress response, thus less performance. This could also explain the differences between treatments in content of the stress responsive c-Myc [SOURCES], meaning higher stress exerted during RT with placebo compared to RT with glucose inducing higher accumulation of c-Myc. If this was the case, it does seem inconclusive that strength did no differ between treatments at earlier or later time points. In addition, there were no differences in training volume per session between treatments, therefore, the mechanical loading should have been equal. Another potential explanation to this observation might be …. (bifasisk restitusjon?)

Despite a thoroughly standardised design, the lack of participants do present a weakness of the present study. Initially, the minimum limit for statistical power was reached however, three participants dropped out during the intervention. Coupled with the fact that no previous study could guide our power calculation, this does introduce a degree of uncertainty regarding our results. Moreover, the study was specifically design to investigate differences in pre-post changes within-participants, and thus also quite limited to this perspective. Hence, the observed effects of time serve as supplementary to the efficiency of the design, without a negative control group to either confirm or refute them. If further investigations are to be made on this topic, these analyses should include a greater sample size, with a negative control. 
In conclusion, resistance exercise training combined with glucose ingestion does not acutely augment ribosome biogenesis following two weeks of heavy resistance training in moderately trained young male and female adults. Neither does it acutely enhance muscular performance or recovery withing 23hrs, measured as unilateral isometric and isokinetic knee extension peak torque.
